### 2015-11-20 paper/1-introduction.tex
Expand/rewrite section1.

%TODO SIZE: approx. 1-2 column, every TODO is approx. one paragraph
%TODO give introduction to the field: parallel computing is necessary with more and more complex systems, blablabla (usual explanation)
%TODO give motivation (adevs is good, but only conservative; exploit new parallelism features of C++11 too)
%TODO give our solution, motivating the need for a different kernel, instead of modifying another kernel (like adevs or PythonPDEVS)

>>>>

%? use paragraphs for layout instead of (dense) flowing text ?
%TODO SIZE: approx. 1-2 column, every TODO is approx. one paragraph
\subsection{DEVS}
%TODO give introduction to the field: DEVS is used for discrete event simulation and is a common basis for most other formalisms
The family of DEVS \cite{DEVSbase} formalisms serve as a common basis for most other discrete event formalisms. Of interest in this paper are the 3 key formalisms: Classic \cite{ClassicDEVS}, Dynamic Structured \cite{DSDEVS} and Parallel \cite{ParallelDEVS} and their implementation. This project uses the DirectConnect \cite{SymbolicFlattening} algorithm, so from a kernel's perspective only Atomic Models exist in the simulation linked to each other by connected ports.

\subsection{Parallel computing}
%TODO parallel computing is necessary with more and more complex systems, blablabla (usual explanation)
Parallel execution of a DEVS simulation can lower runtime and increase the bound on the state space, thereby enabling simulation of more complex systems in the same time-frame.
While the shared memory parallelism offered by most modern hardware does not lower the state space bounds, it can reduce the runtime and offers more direct communication and control between entities involved in synchronization compared to distributed simulation. \\

\subsection{Motivation}
%TODO give motivation (adevs is good, but only conservative; exploit new parallelism features of C++11 too)
Adevs \cite{adevs} offers a very fast conservative synchronized shared memory DEVS simulator, but no optimistic synchronized variant. 
The latter can be significantly faster, especially in simulations where the runtime behaviour of the simulation is hard to predict. \\
The matured parallelism features of C++11 were used to in this project with the dual aim of writing standard-compliant (and thus portable) code without losing access to powerful low-level threading primitives. % may need another pass on this one :(

\subsection{Solution} %for lack of a better word
%TODO give our solution, motivating the need for a different kernel, instead of modifying another kernel (like adevs or PythonPDEVS)
The usage of the DirectConnect\cite{SymbolicFlattening} algorithm makes reusing the adevs kernel hard.
The dxexmachina kernel follows the design of the PythonPDEVS kernel where appropriate, but by the very nature of the implementation languages has to differ in key aspects (e.g. memory allocation strategy). 
The core aims of the project are to offer a deterministic simulation kernel where the simulation author is shielded as much as possible from the kernel implementation, without sacrificing performance. As in PyPDEVS, a model need be written only once for use in the different simulation kernels (with the exception of a non-trivial lookahead()).\\
The tracing framework from PyPDEVS was ported to allow optional verification of simulations. 
\subsection{Time}
The DEVS formalism has $\mathbb{R}$ as time base, but any implementation has to decide on a (finite) representation of time. Dxexmachina kernels can operate on IEEE754 floating point time units, or integral time. The latter significantly reduces the possible range of the simulation, but avoids possible approximation errors in floating point. Furthermore, the notion of $\epsilon$ as the absolute minimum between time points needs to be established, this is non-trivial for floating point. \\
The select() function, which arbitrates between concurrent events, is implicit in this project by providing any time representation with a secondary causality field.

% incomplete

### 2015-11-20 paper/2-background.tex
Expand/rewrite section1.

\subsection{Conservative Synchronization}
%TODO explain conservative synchronization, and how it uses locking to wait for continuing the simulation
Conservative synchronization is defined by the invariant that no model will advance in time before it has received all input from any influencing model. \\
This requires the concepts of eot (earliest output time) and eit (earliest input time) which define the timespan within which a model can safely advance. \\ The eit of any model is the minimum of all eot values of influencing models. A model can simulate up to (but not including) eit, and then has to wait until that value is increased. An important disadvantage here is that the influenced-by relation is always defined at model(link) creation, not at runtime. A model that can influence another, but never does, can severely slow down the protocol.\\ 
Deadlock between models that influence each other can be broken/avoided by a variety of
means, in this simulator the CMB \cite{Chandy:1981:ADS:358598.358613} null-message protocol is used. \\
Null message time is the timestamp a model is guaranteed to have past in simulation. More precisely, a null message of time t is a guarantee that any ouput with timestamp t is already sent.\\
In general, the eot/eit/nulltime of a kernel is the mimimum of each of those values for all models in the kernel.\\
Conservative explicitly relies on information provided by the model creator in the form of lookahead, a relative timespan during which the model is insensitive to outside events. This can be non-trivial to calculate, a simulation writer will in general not be able to predict the exact lookahead of models involved in an experiment without having run the experiment.\\
The performance is more sensitive to lookahead than it is to cycles in the (static) dependency graph.
\subsection{Optimistic Synchronization}
%TODO explain optimistic synchronization, and how it uses rollbacks to undo causality errors
Optimistic synchronization allows causality errors to occur but employs a roll-back mechanism to recover from those errors.
The most common mechanism is the Timewarp \cite{Jefferson:1985:VT:3916.3988} mechanism,
 whenever a kernel receives an event with timestamp in the kernel's past, the state of the kernel is reverted to that time. The gain in runtime this provides is offset by the increase in memory required to keep saved states and (sent) messages.
Optimistic does not rely on any domain specific information, in contrast to conservative. It is only sensitive to runtime use of connections, not the probability that they might occur. %reword connections ?
\\
If the (runtime) dependency graph contains a cycle, optimistic can suffer a series of cascading reverts worsened if the timestamps of the events match exactly. \\
This effect can be lessened by lazy cancellation and/or lazy re-evaluation \cite{FujimotoBook}.
\subsection{Global Virtual Time}
To avoid exhausting memory in state/event saving, optimistic synchronization relies on the concept of global virtual time\cite{Jefferson:1985:VT:3916.3988}. In optimistic simulations, GVT is defined as the lowest timestamp of any unprocessed event. \\ Intuitively this is the timepoint in the simulation that is certain to be preserved, corresponding exactly with the simulation up to that time in a non-parallel implementation.\\
In conservative the minimum simulation time of all kernels is the GVT, or in terms of null messages: the least timestamp of any null message in transit.
In any parallel implementation the GVT calculation is vital to safely commit unrecoverable transactions such as IO (e.g. tracing), releasing memory, ... .

>>>>

\subsection{Conservative Synchronization}
%TODO explain conservative synchronization, and how it uses locking to wait for continuing the simulation
Conservative synchronization is defined by the invariant that no model will advance in time before it has received all input from any influencing model. \\
This requires the concepts of eot (earliest output time) and eit (earliest input time) which define the timespan within which a model can safely advance. \\ The eit of any model is the minimum of all eot values of influencing models. A model can simulate up to (but not including) eit, and then has to wait until that value is increased. An important disadvantage here is that the influenced-by relation is always defined at model(link) creation, not at runtime. A model that can influence another, but never does, can severely slow down the protocol.\\ 
Deadlock between models that influence each other can be broken/avoided by a variety of
means, in this simulator the CMB \cite{Chandy:1981:ADS:358598.358613} null-message protocol is used. \\
Null message time is the timestamp a model is guaranteed to have past in simulation. More precisely, a null message of time t is a guarantee that any ouput with timestamp t-$\epsilon$ is already sent.\\
In general, the eot/eit/nulltime of a kernel is the mimimum of each of those values for all models in the kernel.\\
Conservative explicitly relies on information provided by the model creator in the form of lookahead, a relative timespan during which the model is insensitive to outside events. This can be non-trivial to calculate, a simulation writer will in general not be able to predict the exact lookahead of models involved in an experiment without having run the experiment.\\
The performance is more sensitive to lookahead than it is to cycles in the (static) dependency graph.
\subsection{Optimistic Synchronization}
%TODO explain optimistic synchronization, and how it uses rollbacks to undo causality errors
Optimistic synchronization allows causality errors to occur but employs a roll-back mechanism to recover from those errors.
The most common mechanism is the Timewarp \cite{Jefferson:1985:VT:3916.3988} mechanism,
 whenever a kernel receives an event with timestamp in the kernel's past, the state of the kernel is reverted to that time. The gain in runtime this provides is offset by the increase in memory required to keep saved states and (sent) messages.
Optimistic does not rely on any domain specific information, in contrast to conservative. It is only sensitive to runtime use of connections, not the probability that they might occur. %reword connections ?
\\
If the (runtime) dependency graph contains a cycle, optimistic can suffer a series of cascading reverts worsened if the timestamps of the events match exactly. \\
This effect can be lessened by lazy cancellation and/or lazy re-evaluation \cite{FujimotoBook}.
\subsection{Global Virtual Time}
To avoid exhausting memory in state/event saving, optimistic synchronization relies on the concept of global virtual time\cite{Jefferson:1985:VT:3916.3988}. In optimistic simulations, GVT is defined as the lowest timestamp of any unprocessed event. \\ Intuitively this is the timepoint in the simulation that is certain to be preserved, corresponding exactly with the simulation up to that time in a non-parallel implementation.\\
In conservative the minimum simulation time of all kernels is the GVT, or in terms of null messages: the least timestamp of any null message in transit.
In any parallel implementation the GVT calculation is vital to safely commit unrecoverable transactions such as IO (e.g. tracing), releasing memory, ... .
